---
title: "An Exploration of the Relationship between Air Quality and Solar & Wind Energy Generation in the United States"
subtitle: https://github.com/Am1029/Mujollari_Liao_Mutua_Hoxha/tree/main
author: "Mujollari A., Liao I., Mutua Mulu B., Hoxha L."
output:
  html_document:
    toc: true
    css: styles.css
---

\newpage
\tableofcontents 

```{r Install Packages, warning=FALSE, include=FALSE, echo=FALSE}
# Set your working directory
getwd()

# Load packages
#install.packages('readxl')
#install.packages('foreign')
#install.packages('sf')
#install.packages('leaflet')
#install.packages('leaflet.extras2')
#install.packages('mapview')
#install.packages("httr")
#install.packages("jsonlite") 
#install.packages("gganimate")
#install.packages("https://cran.r-project.org/src/contrib/Archive/transformr/transformr_0.1.3.tar.gz", repos = NULL, type = "source")
#install.packages('animation')
#install.packages('gifski')
#install.packages('gghighlight')
#install.packages('magick')
#install.packages('gtsummary')
#install.packages('janitor')
#install.packages('kableExtra')
```

```{r Loading library, warning=FALSE, include=FALSE, echo=FALSE}
library(readxl)
library(here)
library(tidyverse)
library(lubridate)
library(foreign)
library(dplyr)
library(tidyverse, quietly = TRUE)
library(sf)
library(leaflet)
library(mapview); 
mapviewOptions(fgb = FALSE, basemaps.color.shuffle = FALSE)
library(rvest)
library(httr)
library(jsonlite) 
library(gganimate)
library(transformr)
library(animation)
library(gifski)
library(gghighlight)
library(magick)
library(janitor)
library(gtsummary)
library(leaflet.extras2)
library(kableExtra)
```

```{r Setting Plot Theme, warning=FALSE, include=FALSE, echo=FALSE}
plot_theme<- theme_classic(base_size = 12)+
  theme(
    #plot background
    plot.background = element_rect(color="white"),
    
    #plot boarder
    panel.border = element_blank(),
    
    #plot title
    plot.title=element_text(color="black",hjust=0.5,vjust=1),
    
    #axis labels
    axis.text=element_text(color="black"),
    
    #gridlines 
    panel.grid.major=element_line("lightgray"),
    axis.ticks=element_line(color="white"),
    
    #legend 
    legend.key=element_rect(color="white"),
    legend.background = element_rect(color="white"),
    legend.position="right"
      )

theme_set(plot_theme)
```

```{r Importing Data, warning=FALSE, include=FALSE, echo=FALSE}
# Check working directory
here()

# Load your datasets
#Importing processed energy monthly generation data
df_generation_monthly <- read.csv(here("Data/Processed/df_generation_monthly.csv"),
                                  stringsAsFactors = TRUE) %>% 
  mutate(NetGeneration = NetGeneration/1000000)
#Importing processed pollutant monthly data 
df_pollutant_monthly <- read.csv(here("Data/Processed/df_pollutant.csv"),
                                  stringsAsFactors = TRUE)

#Importing processed plant location and capacity data
df_plantcapacities_locations <- read.csv(here("Data/Processed/df_plantcapacities_locations.csv"),
                                  stringsAsFactors = TRUE)
#Importing processed pollutant observation stations data
df_station_locations <- read.csv(here("Data/Processed/df_latitude_longitude.csv"),
                                  stringsAsFactors = TRUE)
#Importing state abbreviations 
df_state<- read.csv(here("Data/Processed/df_statename.csv"),
                                  stringsAsFactors = TRUE)

#Setting date columns as date objects
df_generation_monthly$Date <- ymd(df_generation_monthly$Date)
df_pollutant_monthly$Date <- ym(df_pollutant_monthly$Date)

#Setting date columns as date objects
df_generation_monthly$Date <- ymd(df_generation_monthly$Date)
df_pollutant_monthly$Date <- ymd(df_pollutant_monthly$Date)

```


# Rationale and Research Questions
## Background 

In the context of escalating global environmental challenges, the shift from traditional fossil fuel-based energy sources to renewable energy has become a focal point in efforts to reduce carbon emissions and combat climate change (Kabeyi et. al, 2022). However, the transition's broader environmental impacts, particularly on air quality, remain less explored. This transition is especially relevant given the increasing global energy demand and the need to meet this demand sustainably.

## Significance 

Understanding the relationship between renewable energy generation and air quality is crucial. Renewable energy sources like wind and solar are lauded for their lower environmental impact compared to fossil fuels, which are major contributors to air pollution (UCSUSA, 2018). Air pollution is a significant environmental hazard, affecting human health, ecosystems, and the climate. It is responsible for millions of premature deaths annually and contributes to the occurrence of diseases like asthma, heart disease, and lung cancer (National Geographic, 2023). Therefore, assessing how increased renewable energy generation affects air quality indicators is not just an environmental concern but a public health imperative.

## Theoretical Context

The hypothesis underlying this research is that an increase in renewable energy generation leads to a reduction in air pollution. This hypothesis is grounded in the understanding that renewable energy sources, unlike fossil fuels, do not emit pollutants like sulfur dioxide, nitrogen oxides, and particulate matter during electricity generation.

## Research Questions

1: What is the relationship between distributed renewable energy generation and the level of air pollution?

This question aims to investigate the correlation between the rise in renewable energy generation and the concentrations of various air pollutants. It seeks to understand whether regions with higher renewable energy output exhibit lower levels of air pollutants.

2: Among air quality indicators (PM2.5, NO2, and SO2), which display the most significant response to variations in energy generation?

This question delves deeper into identifying which specific pollutants are most responsive to changes in energy generation types. It is crucial for pinpointing the environmental benefits of renewable energy sources and for policy-making aimed at targeted air pollution reduction.
\newpage

# Dataset Information

The exploratory analysis required the combination of air quality and power plant datasets. Air quality data in the analysis was obtained from the United States Environmental Protection Agency (EPA) while power plant data was obtained from the U.S. Energy Information Administration (EIA).Our analysis covered a period of two decades, from 2001 to 2021. 

EPA has established air quality standards for six pollutants which include particular matter, ozone, sulfur dioxide, nitrogen dioxide, carbon monoxide, and lead. Among these, particular matter (PM2.5), nitrogen dioxide (NO2), and sulfur dioxide (SO2) are mainly produced as by-products of fossil fuel combustion and pose significant threats to human health (Perera et al. 2019, Perera 2017, Bridges et al. 2015). Thus, this study aims to investigate the relationship between the three air quality indicators, PM2.5, NO2, and SO2, and energy generation. We downloaded the pollutants data from the EPA website that is available from 1980 to 2021 (https://aqs.epa.gov/aqsweb/airdata/download_files.html#Daily). The data is at the county level and consists of number of observation points for each county. We calculated the daily mean by state to retrieve the monthly pollutant concentrations at the state-level.

We collected monthly and annual energy generation from power plants and power plant locations and capacities from two datasets, EIA-923 (https://www.eia.gov/electricity/data/eia923/) and EIA-860 (https://www.eia.gov/electricity/data/eia860/), respectively. The EIA-923 dataset provides us with detailed generator data such as generation, fuel consumption, and stocks. On the other hand, the EIA-860 dataset lists the capacities of power plants, including the generators that are operable, as well as the plant-level data for the surveyed generators, which includes their locations information.

```{r Table: Dataset Information, echo=FALSE, message=FALSE, warning=FALSE}
dataset_info<- data.frame(
  Dataset = c("Air Quality Summary Statistics by Criteria Pollutants and Location","Power Plant Generator Level Capacities and Locations","Power Plant Monthly Energy Generation "),
  Source = c("EPA Air Quality System (AQS) ","EIA Form EIA-860","EIA Form EIA-923"),
  Variables = c("Monthly PM2.5, SO2, and NO2 Concentrations","Annual Installed Generation Capacity by Fuel Type","Monthly Net Generation by Fuel Type"))

info_table<-knitr::kable(dataset_info,caption = "Table 1. Dataset Information for the Sample Period 2001 - 2021")%>%
  kable_styling(full_width = FALSE) #create bottom lines in the table

info_table
```

\newpage

# Exploratory Analysis

We began our exploratory analysis by examining if and how solar and wind energy generating capacity, net generation and air quality has changed over time in the contiguous United States.
We first used the wrangled power plant to visualize the changes in the total installed capacity of solar and wind plants from 2001 to 2021. Figure 1 shows that California, Texas, and Iowa hold the highest cumulative installed capacity, suggesting that these states have significant growth in their installed capacity compared to other states. In addition, we plotted the annual energy generation from solar and wind sources over the same period of time, observing that the states with the highest installed capacity also exhibit the largest annual energy generation from these renewable sources (Figure 2 and Figure 3).

```{r Data Wrangling: Capacity Over Time, echo=FALSE, message=FALSE, warning=FALSE}
#Plotting out trend for top 3 states
top_states_capacity2021 <- df_plantcapacities_locations %>% 
  filter(YEAR == 2021) %>% 
  group_by(STATE) %>% 
  summarise(CAPACITY = sum(CAPACITY)) %>% 
  mutate(Rank = rank(-CAPACITY)) %>% 
  filter(Rank %in% c(1:3))

top_states_renewables2022 <- df_generation_monthly %>% 
  filter(Year == 2022) %>% 
  group_by(State) %>% 
  summarise(NetGeneration = sum(NetGeneration)) %>% 
  mutate(Rank = rank(-NetGeneration)) %>% 
  filter(Rank %in% c(1:3))

#Summarizing monthly data into annual data
df_generation_annual <- df_generation_monthly %>% 
  group_by(State, Year) %>% 
  summarise(NetGeneration = sum(NetGeneration))
```

```{r Visualization: Installed Generation Capacity,echo=FALSE, message=FALSE, warning=FALSE,fig.cap="Figire 1. Annual Installed Generation Capacity: Solar and Wind (MW)", out.width="50%",out.height="50%"}
df_plantcapacities_locations %>%
  group_by(STATE, YEAR) %>% 
  summarise(CAPACITY = sum(CAPACITY)) %>%
  ggplot(aes(x=YEAR,
             y=CAPACITY,
             color=STATE)) +
  geom_line() +
  labs(x="Year",
       y="Installed Capacity (MW)",
       title="Annual Installed Generation Capacity: Solar and Wind (MW)") +
  scale_color_viridis_d()  + #set all states with the same color
  gghighlight(max(CAPACITY)>12000) #only highlight top 3 states with different colors
```

```{r Visualization: Annual Generation,echo=FALSE, message=FALSE, warning=FALSE,fig.cap="Figire 2. Annual Solar and Wind Energy Generation over time (TWh)", out.width="50%",out.height="50%"}
df_generation_annual %>% 
  ggplot(aes(x=Year,
             y=NetGeneration,
             color=State)) +
  geom_line() +
  gghighlight(max(NetGeneration)>40) +
  labs(x="Year",
       y="Annual Energy Generation (TWh)",
       title="Annual Solar and Wind Energy Generation over time (TWh)") +
  scale_color_viridis_d() 
```

```{r Visualization: Monthly Generation,echo=FALSE, message=FALSE, warning=FALSE,fig.cap="Figire 3. Monthly Renewable Energy Generation per State (TWh)", out.width="50%",out.height="50%"}
df_generation_monthly %>%
  ggplot(aes(x=Date,
             y=NetGeneration,
             color=State)) +
  geom_line() +
  gghighlight(max(NetGeneration)>4.5) +
  labs(x="Date",
       y="Monthly Energy Generation (TWh)",
       title="Monthly Renewable Energy Generation per State (TWh)") +
  scale_color_viridis_d()
```
  
The change in installed solar and wind plants can be also be visualized spatially across the contiguous United States as illustrated below. From the map, we note that the number of installed solar and wind plants increased significantly over the two decades from 2001 to 2021.

```{r Visualization: Exploratory Spatial, echo=FALSE, message=FALSE, warning=FALSE,fig.cap="Figure 4. Solar and Wind Power Plants Locations", out.width="100%"}

df_plantcapacities_locations_sf_2001 <- df_plantcapacities_locations %>%
  na.omit() %>%
  st_as_sf(coords = c('LONG','LAT'), crs=4269) %>% 
  filter(YEAR == 2001)
df_plantcapacities_locations_sf_2021 <- df_plantcapacities_locations %>%
  na.omit() %>%
  st_as_sf(coords = c('LONG','LAT'), crs=4269) %>% 
  filter(YEAR == 2021)

mapview(df_plantcapacities_locations_sf_2001,
        cex = 'CAPACITY',
        col.regions = '#2b83ba',
        layer.name = '2001') |
  mapview(df_plantcapacities_locations_sf_2021,
        cex = 'CAPACITY',
        col.regions = '#f7fcb9',
        layer.name = '2021')
```

We utilized ggplot and gganimate to visualize the increase and distribution of plants within the states that had the highest growth in installed capacity. We thus found the top three states and created a visualization that is shown below:

```{r Visualization: Top 3 States, echo=FALSE, message=FALSE, warning=FALSE}
# Plot state boundary
#CA
state.sf.CA<- st_read(here('Data/Spatial/cb_2018_us_county_20m.shp'),
                      quiet = TRUE) %>% 
  filter(STATEFP == '06')
#TX
state.sf.TX<- st_read(here('Data/Spatial/cb_2018_us_county_20m.shp'),
                      quiet = TRUE) %>% 
  filter(STATEFP == '48')
#IA
state.sf.IA<- st_read(here('Data/Spatial/cb_2018_us_county_20m.shp'),
                      quiet = TRUE) %>% 
  filter(STATEFP == '19')

# Map out power plant location over time
#CA
df_plantcapacities_locations_sf_CA <- df_plantcapacities_locations %>%
  na.omit() %>%
  st_as_sf(coords = c('LONG','LAT'), crs=4269) %>% 
  mutate(Date = ymd(YEAR, truncated = 2L)) %>% 
  filter(STATE == 'CA')

animated_map_CA <- ggplot() +
  geom_sf(data=state.sf.CA, fill='#addd8e') +
  geom_sf(data=df_plantcapacities_locations_sf_CA, color = '#f7fcb9', aes(size = CAPACITY)) +
  labs(title = 'Year: {frame_time}',
       subtitle = "California Solar and Wind Installed Capacity") +
  coord_sf(crs = 4269, datum = NA) +
  transition_time(YEAR) +
  enter_fade() +
  exit_fade() +
  shadow_mark()

#TX
df_plantcapacities_locations_sf_TX <- df_plantcapacities_locations %>%
  na.omit() %>%
  st_as_sf(coords = c('LONG','LAT'), crs=4269) %>% 
  mutate(Date = ymd(YEAR, truncated = 2L)) %>% 
  filter(STATE == 'TX')

animated_map_TX <- ggplot() +
  geom_sf(data=state.sf.TX, fill='#addd8e') +
  geom_sf(data=df_plantcapacities_locations_sf_TX, color = '#f7fcb9', aes(size = CAPACITY)) +
  labs(title = 'Year: {frame_time}',
       subtitle = "Texas Solar and Wind Installed Capacity") +
  coord_sf(crs = 4269, datum = NA) +
  transition_time(YEAR) +
  enter_fade() +
  exit_fade() +
  shadow_mark()

#IA
df_plantcapacities_locations_sf_IA <- df_plantcapacities_locations %>%
  na.omit() %>%
  st_as_sf(coords = c('LONG','LAT'), crs=4269) %>% 
  mutate(Date = ymd(YEAR, truncated = 2L)) %>% 
  filter(STATE == 'IA')

animated_map_IA <- ggplot() +
  geom_sf(data=state.sf.IA, fill='#addd8e') +
  geom_sf(data=df_plantcapacities_locations_sf_IA, color = '#f7fcb9', aes(size = CAPACITY)) +
  labs(title = 'Year: {frame_time}',
       subtitle = "Iowa Solar and Wind Installed Capacity") +
  coord_sf(crs = 4269, datum = NA) +
  transition_time(YEAR) +
  enter_fade() +
  exit_fade() +
  shadow_mark()
```

```{r Create GIF, echo=FALSE, message=FALSE, warning=FALSE,fig.show="hold",fig.cap="Figure. Solar and Wind Insatlled Capacities over Time"}

#CA_gif <- animate(animated_map_CA, renderer = gifski_renderer(), fps = 20, duration = 30) 
#TX_gif <- animate(animated_map_TX, renderer = gifski_renderer(), fps = 20, duration = 30)
#IA_gif <- animate(animated_map_IA, renderer = gifski_renderer(), fps = 20, duration = 30)
 

# Create GIF
#CA_gif <- animate(animated_map_CA, width = 300, height = 300) 
#TX_gif <- animate(animated_map_TX, width = 300, height = 300)
#IA_gif <- animate(animated_map_IA, width = 300, height = 300)
#notes: still have error message

#CA_mgif <- image_read(CA_gif)
#TX_mgif <- image_read(TX_gif)
#IA_mgif <- image_read(IA_gif)

#states_gif <- image_append(c(CA_mgif[1], TX_mgif[1], IA_mgif[1])) #combine 3 states gifs into 1 gif

#for(i in 2:96) {
#  combined <- image_append(c(CA_mgif[i], TX_mgif[i], IA_mgif[i]))
#  states_gif <- c(states_gif, combined)}

#states_gif
```

After analyzing the increasing installed capacity of solar and wind plants in the United States, with a special focus on California, Texas, and Iowa, we proceeded to determine the concentration changes of critical criteria pollutants over time. To simplify the process, we used visualizations to to present the data on the reduction of SO2, NO2, and PM2.5 over time, which are associated with fossil fuel generation. As per our findings below, it's evident that the concentration of these pollutants has been steadily decreasing over the years.

```{r Pollutant Data Wrangling, echo=FALSE, message=FALSE, warning=FALSE}
df_pollutant_location<-merge(df_pollutant_monthly,df_state,by='State') #add state abbreviation

# Select air quality variation among the top 3 states in renewable energy generation
df_pollutant_top3<-df_pollutant_location %>%
  filter(Abbreviation %in% top_states_renewables2022$State) %>% 
  arrange(Date)

#CO
#df_CO_top3<-df_pollutant_top3 %>%
#  filter(Pollutant=="Carbon monoxide")%>% 
#  mutate(Year = year(Date)) %>% 
#  mutate(Year = as.factor(Year))

#PM2.5
df_PM2.5_top3<-df_pollutant_top3 %>%
  filter(Pollutant=="PM2.5 - Local Conditions") %>% 
  mutate(Year = year(Date)) %>% 
  mutate(Year = as.factor(Year))

#SO2
df_SO2_top3<-df_pollutant_top3 %>%
  filter(Pollutant=="Sulfur dioxide") %>% 
  mutate(Year = year(Date)) %>% 
  mutate(Year = as.factor(Year))

#NO2
df_NO2_top3<-df_pollutant_top3 %>%
  filter(Pollutant=="Nitrogen dioxide (NO2)") %>% 
  mutate(Year = year(Date)) %>% 
  mutate(Year = as.factor(Year))

#PM10
#df_PM10_top10<-df_pollutant_top10 %>%
#  filter(Pollutant=="PM10 Total 0-10um STP") %>% 
#  mutate(Year = year(Date)) %>% 
#  mutate(Year = as.factor(Year))
```


```{r Visualization Pollutants, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold",fig.cap="Trends in Pollutant Concentrations", out.width="50%"}
ggplot(df_SO2_top3, aes(x =Date, y = Mean, color=State)) +
  geom_line()+
  labs(x="Date",y="Parts per million")+
  ggtitle("Trends in SO2 Concentrations")+
  theme(plot.title = element_text(hjust = 0.5))

ggplot(df_PM2.5_top3, aes(x =Date, y = Mean, color=State)) +
  geom_line()+
  labs(x="Date",y="mg/m\u00B3")+
  ggtitle("Trends in PM2.5 Concentrations")+
  theme(plot.title = element_text(hjust = 0.5))

ggplot(df_NO2_top3, aes(x =Date, y = Mean, color=State)) +
  geom_line()+
  labs(x="Date",y="Parts per billion")+
  ggtitle("Trends in NO2 Concentrations")+
  theme(plot.title = element_text(hjust = 0.5))
```
\newpage
```{r Variable Visualization Function, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold", out.width="50%"}

visualizeVariables <- function(selected_state, selected_pollutant){
  #Filtering out energy data for the selected state
  df_energy.data <- df_generation_monthly %>% 
  filter(State == selected_state) %>%  #Filtering out California
  filter(Year  %in% c(2001:2021))
  
  if(selected_pollutant == 'PM2.5 - Local Conditions'){
    df_energy.air.data <- df_pollutant_top3 %>%
      filter(Abbreviation == selected_state) %>% #Filtering out selected state
      filter(Pollutant == selected_pollutant) %>%
      select(Date, Mean, Abbreviation) %>% #Filtering out:`Date`, `Mean`, `Abbreviation`
      rename(MeanPM25 = Mean, State = Abbreviation) %>% #Renaming columns
      left_join(df_energy.data, by = c('Date', 'State')) #Combining air quality and energy dataframes
    
    #Plotting the regression
  plot_energy.air <- ggplot(df_energy.air.data, aes(x = NetGeneration, y = MeanPM25)) +
    geom_point() +
    geom_smooth(method = "lm", color = "red") +
    labs(x="Net Generation (TWh)", 
       y="Mean PM2.5 (µg/m³)",
       title= paste0("Mean PM2.5 (µg/m³) by Net Generation (TWh) in ",
                     df_state$State[df_state$Abbreviation == selected_state]))
  return(list(df_energy.air.data, plot_energy.air))}
  
  else if(selected_pollutant == 'Sulfur dioxide'){
    df_energy.air.data <- df_pollutant_top3 %>%
      filter(Abbreviation == selected_state) %>% #Filtering out selected state
      filter(Pollutant == selected_pollutant) %>%
      select(Date, Mean, Abbreviation) %>% #Filtering out:`Date`, `Mean`, `Abbreviation`
      rename(MeanSO2 = Mean, State = Abbreviation) %>% #Renaming columns
      left_join(df_energy.data, by = c('Date', 'State')) #Combining air quality and energy dataframes
    
    #Plotting the regression
    plot_energy.air <- ggplot(df_energy.air.data, aes(x = NetGeneration, y = MeanSO2)) +
      geom_point() +
      geom_smooth(method = "lm", color = "red") +
      labs(x="Net Generation (TWh)",
           y="SO2 Concentration (ppb)",
           title= paste0("SO2 Concentration (ppb) by Net Generation (TWh) in ", 
                         df_state$State[df_state$Abbreviation == selected_state]))
    
    return(list(df_energy.air.data, plot_energy.air))
    
  }else if(selected_pollutant == 'Nitrogen dioxide (NO2)'){
    df_energy.air.data <- df_pollutant_top3 %>%
      filter(Abbreviation == selected_state) %>% #Filtering out selected state
      filter(Pollutant == selected_pollutant) %>%
      select(Date, Mean, Abbreviation) %>% #Filtering out:`Date`, `Mean`, `Abbreviation`
      rename(MeanNO2 = Mean, State = Abbreviation) %>% #Renaming columns
      left_join(df_energy.data, by = c('Date', 'State')) #Combining air quality and energy dataframes
    
    #Plotting the regression
    plot_energy.air <- ggplot(df_energy.air.data, aes(x = NetGeneration, y = MeanNO2)) +
      geom_point() +
      geom_smooth(method = "lm", color = "red") +
      labs(x="Net Generation (TWh)",
           y="Nitrogen Dioxide Concentration (µg/m³)",
           title= paste0("Nitrogen Dioxide Concentration (µg/m³) by Net Generation (TWh)
                         in ",
                         df_state$State[df_state$Abbreviation == selected_state]))
    return(list(df_energy.air.data, plot_energy.air))
        
  }
  
}

df_CA.energy.air.data.PM2.5 <- visualizeVariables('CA','PM2.5 - Local Conditions')[[1]]

df_TX.energy.air.data.PM2.5 <- visualizeVariables('TX','PM2.5 - Local Conditions')[[1]]

df_IA.energy.air.data.PM2.5 <- visualizeVariables('IA','PM2.5 - Local Conditions')[[1]]

df_CA.energy.air.data.SO2 <- visualizeVariables('CA','Sulfur dioxide')[[1]]

df_TX.energy.air.data.SO2 <- visualizeVariables('TX','Sulfur dioxide')[[1]]

df_IA.energy.air.data.SO2 <- visualizeVariables('IA','Sulfur dioxide')[[1]]

df_CA.energy.air.data.NOX <- visualizeVariables('CA','Nitrogen dioxide (NO2)')[[1]]

df_TX.energy.air.data.NOX <- visualizeVariables('TX','Nitrogen dioxide (NO2)')[[1]]

df_IA.energy.air.data.NOX <- visualizeVariables('IA','Nitrogen dioxide (NO2)')[[1]]

```
  
To investigate the trend in the measured pollutants over time, we conducted a time series analysis of the measured values of each of the three pollutants in California, Texas, and Iowa from 2001 to 2021. Our goal was to determine whether there has been a change in the recorded PM2.5, SO2, and NO2 concentrations over the sample period. The null hypothesis is that there has been no change in the recorded PM2.5, SO2 and NO2 concentrations in the three states over the sample period. The alternative hypothesis is that there has been a change in the recorded PM2.5, SO2 and NO2 concentrations in the three states over the sample period.   

After decomposing the time series, we observed that each of the three pollutants has a seasonal component as observed in the time series plots shown below. Hence we run the seasonal Mann-Kendall (SMK) test on each dataset, which produced a p-value of less than 0.05 (<0.05) for each time series. As a result, we can reject the null hypothesis, and the analysis indicates that there has been a change in the recorded PM2.5, SO2, and NO2 concentrations over the period 2001 - 2021 in California, Texas, and Iowa. The negative tau values suggests a negative correlation which implies that the change for each pollutant in each state has been a decrease.  

```{r Time Series, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold", out.width="50%"}
#Generating PM2.5 time series objects
first_month <- month(first(df_CA.energy.air.data.PM2.5$Date))
first_year <- year(first(df_CA.energy.air.data.PM2.5$Date))

CA.PM2.5.monthly.ts <- ts(df_CA.energy.air.data.PM2.5$Mean,
                        start=c(first_year,first_month),
                        frequency=12) 
CA.SO2.monthly.ts <- ts(df_CA.energy.air.data.SO2$Mean,
                        start=c(first_year,first_month),
                        frequency=12) 
CA.NOX.monthly.ts <- ts(df_CA.energy.air.data.NOX$Mean,
                        start=c(first_year,first_month),
                        frequency=12) 
TX.PM2.5.monthly.ts <- ts(df_TX.energy.air.data.PM2.5$Mean,
                        start=c(first_year,first_month),
                        frequency=12) 
TX.SO2.monthly.ts <- ts(df_TX.energy.air.data.SO2$Mean,
                        start=c(first_year,first_month),
                        frequency=12) 
TX.NOX.monthly.ts <- ts(df_TX.energy.air.data.NOX$Mean,
                        start=c(first_year,first_month),
                        frequency=12) 
IA.PM2.5.monthly.ts <- ts(df_IA.energy.air.data.PM2.5$Mean,
                        start=c(first_year,first_month),
                        frequency=12) 
IA.SO2.monthly.ts <- ts(df_IA.energy.air.data.SO2$Mean,
                        start=c(first_year,first_month),
                        frequency=12) 
IA.NOX.monthly.ts <- ts(df_IA.energy.air.data.NOX$Mean,
                        start=c(first_year,first_month),
                        frequency=12) 

#Decomposing the monthly time series objects and plotting
CA.PM2.5.monthly.data.decomp <- stl(CA.PM2.5.monthly.ts,s.window = "periodic")
CA.SO2.monthly.data.decomp <- stl(CA.SO2.monthly.ts,s.window = "periodic")
CA.NOX.monthly.data.decomp <- stl(CA.NOX.monthly.ts,s.window = "periodic")
TX.PM2.5.monthly.data.decomp <- stl(TX.PM2.5.monthly.ts,s.window = "periodic")
TX.SO2.monthly.data.decomp <- stl(TX.SO2.monthly.ts,s.window = "periodic")
TX.NOX.monthly.data.decomp <- stl(TX.NOX.monthly.ts,s.window = "periodic")
IA.PM2.5.monthly.data.decomp <- stl(IA.PM2.5.monthly.ts,s.window = "periodic")
IA.SO2.monthly.data.decomp <- stl(IA.SO2.monthly.ts,s.window = "periodic")
IA.NOX.monthly.data.decomp <- stl(IA.NOX.monthly.ts,s.window = "periodic")

plot(CA.PM2.5.monthly.data.decomp, main = "PM2.5 in California")
plot(CA.SO2.monthly.data.decomp, main = "SO2 in California")
plot(CA.NOX.monthly.data.decomp, main = "NO2 in California")
plot(TX.PM2.5.monthly.data.decomp, main = "PM2.5 in Texas")
plot(TX.SO2.monthly.data.decomp, main = "SO2 in Texas")
plot(TX.NOX.monthly.data.decomp, main = "NO2 in Texas")
plot(IA.PM2.5.monthly.data.decomp, main = "PM2.5 in Iowa")
plot(IA.SO2.monthly.data.decomp, main = "SO2 in Iowa")
plot(IA.NOX.monthly.data.decomp, main = "NO2 in Iowa")

#Running monotonic trend analysis
# Running the seasonal Mann-Kendall test
CA.PM2.5.data.trend <- Kendall::SeasonalMannKendall(CA.PM2.5.monthly.ts)
CA.SO2.data.trend <- Kendall::SeasonalMannKendall(CA.SO2.monthly.ts)
CA.NOX.data.trend <- Kendall::SeasonalMannKendall(CA.NOX.monthly.ts)
TX.PM2.5.data.trend <- Kendall::SeasonalMannKendall(TX.PM2.5.monthly.ts)
TX.SO2.data.trend <- Kendall::SeasonalMannKendall(TX.SO2.monthly.ts)
TX.NOX.data.trend <- Kendall::SeasonalMannKendall(TX.NOX.monthly.ts)
IA.PM2.5.data.trend <- Kendall::SeasonalMannKendall(IA.PM2.5.monthly.ts)
IA.SO2.data.trend <- Kendall::SeasonalMannKendall(IA.SO2.monthly.ts)
IA.NOX.data.trend <- Kendall::SeasonalMannKendall(IA.NOX.monthly.ts)

# Inspecting results
CA_PM2.5 <- data.frame(t(unlist(CA.PM2.5.data.trend)))
CA_SO2 <- data.frame(t(unlist(CA.SO2.data.trend)))
CA_NOX <- data.frame(t(unlist(CA.NOX.data.trend)))
TX_PM2.5 <- data.frame(t(unlist(TX.PM2.5.data.trend)))
TX_SO2 <- data.frame(t(unlist(TX.SO2.data.trend)))
TX_NOX <- data.frame(t(unlist(TX.NOX.data.trend)))
IA_PM2.5 <- data.frame(t(unlist(IA.PM2.5.data.trend)))
IA_SO2 <- data.frame(t(unlist(IA.SO2.data.trend)))
IA_NOX <- data.frame(t(unlist(IA.NOX.data.trend)))


SMKResults <- bind_rows(lst(CA_PM2.5,
                        CA_SO2,
                        CA_NOX,
                        TX_PM2.5,
                        TX_SO2,
                        TX_NOX,
                        IA_PM2.5,
                        IA_SO2,
                        IA_NOX),
                        .id = "id") %>% 
  select(id, tau, sl)

SMKResultsTable <- knitr::kable(SMKResults,
             col.names = c("Trend","tau","2-sided pvalue"),
             caption = "Results of Seasonal Mann-Kendall test") %>% 
  kable_styling(full_width = FALSE) #create bottom lines in the table

SMKResultsTable
```

# Analysis
## Question 1: 
What is the relationship between distributed renewable energy generation and the level of air pollution?

We can formulate a null and alternative hypothesis for the above research question as follows:  

>H0: There is no change in recorded air quality with an increase in renewable energy generation in the states of California, Texas and Iowa over the period 2001 - 2021.  
>Ha: There is a change in recorded air quality with an increase in renewable energy generation in the states of California, Texas and Iowa over the period 2001 - 2021.

To evaluate this hypothesis, we generated a plot of Mean PM2.5 measured against net monthly solar and wind generation for all three states.

```{r variableVisualizationPlots, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold", out.width="50%"}

print(visualizeVariables('CA','PM2.5 - Local Conditions')[[2]])

print(visualizeVariables('TX','PM2.5 - Local Conditions')[[2]])

print(visualizeVariables('IA','PM2.5 - Local Conditions')[[2]])

print(visualizeVariables('CA','Sulfur dioxide')[[2]])

print(visualizeVariables('TX','Sulfur dioxide')[[2]])

print(visualizeVariables('IA','Sulfur dioxide')[[2]])

print(visualizeVariables('CA','Nitrogen dioxide (NO2)')[[2]])

print(visualizeVariables('TX','Nitrogen dioxide (NO2)')[[2]])

print(visualizeVariables('IA','Nitrogen dioxide (NO2)')[[2]])

```

The figures above suggest that the measured value of the pollutants has an inverse relationship or negative correlation with net generation from solar and wind. This implies that the higher the amount of energy generated from wind and solar, the lower the amount of the three criteria pollutants. To investigate this further, we performed a simple linear regression of the relationship between the mean quantity of each pollutant and net energy generation with the results summarized in the table outlined below:  

### Simple Linear Regression Results
```{r SimpleLinearRegression, echo=FALSE, message=FALSE, warning=FALSE}
PM25.CA.regression <- lm(
  data = df_CA.energy.air.data.PM2.5,
  MeanPM25 ~ NetGeneration)

SO2.CA.regression <- lm(
  data = df_CA.energy.air.data.SO2,
  MeanSO2 ~ NetGeneration)

NOX.CA.regression <- lm(
  data = df_CA.energy.air.data.NOX,
  MeanNO2 ~ NetGeneration)

CA.t1 <- tbl_regression(PM25.CA.regression,
                        label = list(NetGeneration ~ "Net Generation")) %>% 
  bold_p() %>% 
  add_glance_table(
    label = list(sigma ~ "\U03C3"),
    include = c(r.squared, AIC, sigma)
  )
CA.t2 <- tbl_regression(SO2.CA.regression,
                        label = list(NetGeneration ~ "Net Generation")) %>% 
  bold_p() %>% 
  add_glance_table(
    label = list(sigma ~ "\U03C3"),
    include = c(r.squared, AIC, sigma)
  )
CA.t3 <- tbl_regression(NOX.CA.regression,
                        label = list(NetGeneration ~ "Net Generation")) %>% 
  bold_p() %>% 
  add_glance_table(
    label = list(sigma ~ "\U03C3"),
    include = c(r.squared, AIC, sigma)
  )

tbl_merge_CA <-
  tbl_merge(
    tbls = list(CA.t1, CA.t2, CA.t3),
    tab_spanner = c("**PM2.5**", "**SO2**", "**NO2**")
  )

PM25.TX.regression <- lm(
  data = df_TX.energy.air.data.PM2.5,
  MeanPM25 ~ NetGeneration)

SO2.TX.regression <- lm(
  data = df_TX.energy.air.data.SO2,
  MeanSO2 ~ NetGeneration)

NOX.TX.regression <- lm(
  data = df_TX.energy.air.data.NOX,
  MeanNO2 ~ NetGeneration)


TX.t1 <- tbl_regression(PM25.TX.regression,
                        label = list(NetGeneration ~ "Net Generation")) %>% 
  bold_p() %>% 
  add_glance_table(
    label = list(sigma ~ "\U03C3"),
    include = c(r.squared, AIC, sigma)
  )
TX.t2 <- tbl_regression(SO2.TX.regression,
                        label = list(NetGeneration ~ "Net Generation")) %>%
  bold_p() %>% 
  add_glance_table(
    label = list(sigma ~ "\U03C3"),
    include = c(r.squared, AIC, sigma)
  )
TX.t3 <- tbl_regression(NOX.TX.regression,
                        label = list(NetGeneration ~ "Net Generation")) %>%
  bold_p() %>% 
  add_glance_table(
    label = list(sigma ~ "\U03C3"),
    include = c(r.squared, AIC, sigma)
  )

tbl_merge_TX <-
  tbl_merge(
    tbls = list(TX.t1, TX.t2, TX.t3),
    tab_spanner = c("**PM2.5**", "**SO2**", "**NO2**")
  )


PM25.IA.regression <- lm(
  data = df_IA.energy.air.data.PM2.5,
  MeanPM25 ~ NetGeneration)

SO2.IA.regression <- lm(
  data = df_IA.energy.air.data.SO2,
  MeanSO2 ~ NetGeneration)

NOX.IA.regression <- lm(
  data = df_IA.energy.air.data.NOX,
  MeanNO2 ~ NetGeneration)

IA.t1 <- tbl_regression(PM25.IA.regression,
                        label = list(NetGeneration ~ "Net Generation")) %>%
  bold_p() %>% 
  add_glance_table(
    label = list(sigma ~ "\U03C3"),
    include = c(r.squared, AIC, sigma)
  )
IA.t2 <- tbl_regression(SO2.IA.regression,
                        label = list(NetGeneration ~ "Net Generation")) %>%
  bold_p() %>% 
  add_glance_table(
    label = list(sigma ~ "\U03C3"),
    include = c(r.squared, AIC, sigma)
  )
IA.t3 <- tbl_regression(NOX.IA.regression,
                        label = list(NetGeneration ~ "Net Generation")) %>%
  bold_p() %>% 
  add_glance_table(
    label = list(sigma ~ "\U03C3"),
    include = c(r.squared, AIC, sigma)
  )


tbl_merge_IA <-
  tbl_merge(
    tbls = list(IA.t1, IA.t2, IA.t3),
    tab_spanner = c("**PM2.5**", "**SO2**", "**NO2**")
  ) 

tbl_stack_regressionTables <-
  tbl_stack(list(tbl_merge_CA, tbl_merge_TX, tbl_merge_IA),
            group_header = c("California", "Texas", "Iowa"))

tbl_stack_regressionTables

```
  
#Interpretation (to be edited)

PM2.5 and Net Generation:   

The beta coefficient of net renewable energy generation which is the slope of the regression line estimating the relationship between PM2.5 concentration and renewable energy generation is -0.77, -0.28 and -1.3 respectively for California, Texas and Iowa. This implies that an increase in solar and wind energy generation is resulting in a decrease in concentrations of PM2.5. Assuming a confidence interval of 0.05, the p values are less than our confidence level which implies that the results are statistically significant and there is a significant negative correlation between PM2.5 concentration and renewable energy generation. Based on the R-squared values, 8.8%, 17.7% and 21.4% of the total variance in PM2.5 concentration in California, Texas and Iowa respectively can be explained by renewable energy generation.

Sulfur Dioxide (SO2) and Net Generation:   

The beta coefficient of net renewable energy generation which is the slope of the regression line estimating the relationship between SO2 concentration and renewable energy generation is -0.21, -0.09 and -0.64 respectively for California, Texas and Iowa. This implies that an increase in solar and wind energy generation is resulting in a decrease in concentrations of SO2. Assuming a confidence interval of 0.05, the p values are less than our confidence level which implies that the results are statistically significant and there is a significant negative correlation between SO2 concentration and renewable energy generation. Based on the R-squared values, 44.5%, 21.3% and 42.3% of the total variance in SO2 concentration in California, Texas and Iowa respectively can be explained by renewable energy generation.

Nitrogen Dioxide (NO2) and Net Generation:   

The beta coefficient of net renewable energy generation which is the slope of the regression line estimating the relationship between NO2 concentration and renewable energy generation is -1.8, -0.47 and -1.5 respectively for California, Texas and Iowa. This implies that an increase in solar and wind energy generation is resulting in a decrease in concentrations of NO2. Assuming a confidence interval of 0.05, the p values are less than our confidence level which implies that the results are statistically significant and there is a significant negative correlation between NO2 concentration and renewable energy generation. Based on the R-squared values, 46.4%, 25.5% and 35.2% of the total variance in NO2 concentration in California, Texas and Iowa respectively can be explained by renewable energy generation.


So, in conclusion, these observations suggest an inverse relationship where increased renewable energy generation could be associated with decreased emissions of PM2.5, SO2 and NO2, which are pollutants typically associated with the combustion of fossil fuels.  

  
  
### Multiple Linear Regression Results

```{r MultipleLinearRegression, echo=FALSE, message=FALSE, warning=FALSE}

PM25.CA.regression <- lm(
  data = df_CA.energy.air.data.PM2.5,
  MeanPM25 ~ NetGeneration + Month + Year)

SO2.CA.regression <- lm(
  data = df_CA.energy.air.data.SO2,
  MeanSO2 ~ NetGeneration + Month + Year)

NOX.CA.regression <- lm(
  data = df_CA.energy.air.data.NOX,
  MeanNO2 ~ NetGeneration + Month + Year)

CA.t1 <- tbl_regression(PM25.CA.regression,
                        label = list(NetGeneration ~ "Net Generation")) %>% 
  bold_p() %>% 
  add_glance_table(
    label = list(sigma ~ "\U03C3"),
    include = c(r.squared, AIC, sigma)
  )
CA.t2 <- tbl_regression(SO2.CA.regression,
                        label = list(NetGeneration ~ "Net Generation")) %>% 
  bold_p() %>% 
  add_glance_table(
    label = list(sigma ~ "\U03C3"),
    include = c(r.squared, AIC, sigma)
  )
CA.t3 <- tbl_regression(NOX.CA.regression,
                        label = list(NetGeneration ~ "Net Generation")) %>% 
  bold_p() %>% 
  add_glance_table(
    label = list(sigma ~ "\U03C3"),
    include = c(r.squared, AIC, sigma)
  )

tbl_merge_CA <-
  tbl_merge(
    tbls = list(CA.t1, CA.t2, CA.t3),
    tab_spanner = c("**PM2.5**", "**SO2**", "**NO2**")
  )

PM25.TX.regression <- lm(
  data = df_TX.energy.air.data.PM2.5,
  MeanPM25 ~ NetGeneration + Month + Year)

SO2.TX.regression <- lm(
  data = df_TX.energy.air.data.SO2,
  MeanSO2 ~ NetGeneration + Month + Year)

NOX.TX.regression <- lm(
  data = df_TX.energy.air.data.NOX,
  MeanNO2 ~ NetGeneration + Month + Year)


TX.t1 <- tbl_regression(PM25.TX.regression,
                        label = list(NetGeneration ~ "Net Generation")) %>% 
  bold_p() %>% 
  add_glance_table(
    label = list(sigma ~ "\U03C3"),
    include = c(r.squared, AIC, sigma)
  )
TX.t2 <- tbl_regression(SO2.TX.regression,
                        label = list(NetGeneration ~ "Net Generation")) %>%
  bold_p() %>% 
  add_glance_table(
    label = list(sigma ~ "\U03C3"),
    include = c(r.squared, AIC, sigma)
  )
TX.t3 <- tbl_regression(NOX.TX.regression,
                        label = list(NetGeneration ~ "Net Generation")) %>%
  bold_p() %>% 
  add_glance_table(
    label = list(sigma ~ "\U03C3"),
    include = c(r.squared, AIC, sigma)
  )

tbl_merge_TX <-
  tbl_merge(
    tbls = list(TX.t1, TX.t2, TX.t3),
    tab_spanner = c("**PM2.5**", "**SO2**", "**NO2**")
  )


PM25.IA.regression <- lm(
  data = df_IA.energy.air.data.PM2.5,
  MeanPM25 ~ NetGeneration + Month + Year)

SO2.IA.regression <- lm(
  data = df_IA.energy.air.data.SO2,
  MeanSO2 ~ NetGeneration + Month + Year)

NOX.IA.regression <- lm(
  data = df_IA.energy.air.data.NOX,
  MeanNO2 ~ NetGeneration + Month + Year)

IA.t1 <- tbl_regression(PM25.IA.regression,
                        label = list(NetGeneration ~ "Net Generation")) %>%
  bold_p() %>% 
  add_glance_table(
    label = list(sigma ~ "\U03C3"),
    include = c(r.squared, AIC, sigma)
  )
IA.t2 <- tbl_regression(SO2.IA.regression,
                        label = list(NetGeneration ~ "Net Generation")) %>%
  bold_p() %>% 
  add_glance_table(
    label = list(sigma ~ "\U03C3"),
    include = c(r.squared, AIC, sigma)
  )
IA.t3 <- tbl_regression(NOX.IA.regression,
                        label = list(NetGeneration ~ "Net Generation")) %>%
  bold_p() %>% 
  add_glance_table(
    label = list(sigma ~ "\U03C3"),
    include = c(r.squared, AIC, sigma)
  )


tbl_merge_IA <-
  tbl_merge(
    tbls = list(IA.t1, IA.t2, IA.t3),
    tab_spanner = c("**PM2.5**", "**SO2**", "**NO2**")
  ) 

tbl_stack_regressionTables <-
  tbl_stack(list(tbl_merge_CA, tbl_merge_TX, tbl_merge_IA),
            group_header = c("California", "Texas", "Iowa"))

tbl_stack_regressionTables

```
  
## Question 2: 
Among three air quality indicators (PM2.5, NO2, and SO2), which display the most significant response to variations in energy generation?

We can formulate a null and alternative hypothesis for the above research question as follows:  
  
>H0: There has been a uniform impact on all three air quality indicator by an increase in renewable energy generation in the states of California, Texas and Iowa over the period 2001 - 2021.  
>Ha: There has not been a uniform impact on all three air quality indicator by an increase in renewable energy generation in the states of California, Texas and Iowa over the period 2001 - 2021.

To evaluate the above research question, we plotted the distribution of quantity of the pollutants over time as demonstrated in the box plots below.  


```{r Boxplots, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 12}
#Box Plot for Comparing Distributions of a Single Pollutant Across Multiple States:


# PM2.5
ggplot(df_PM2.5_top3, aes(x = Year,
                          y = Mean,
                          color = State)) +
  geom_boxplot() +
  labs(title = "Distribution of PM2.5 Levels Across States",
       x = "Year",
       y = "PM2.5 Concentration (µg/m³)",
       color = "State") +
  theme(legend.position = "bottom")

# CO

#ggplot(df_CO_top3, aes(x = State, y = Mean, fill = State)) +
#  geom_boxplot() +
#  labs(title = "Distribution of CO Levels Across States",
#       x = "State",
#       y = "CO Concentration (ppm)") +
#  theme_minimal() +
#  theme(legend.position = "none")

# NO2

ggplot(df_NO2_top3, aes(x = Year,
                        y = Mean,
                        color = State)) +
  geom_boxplot() +
  labs(title = "Distribution of NO2 Levels Across States",
       x = "Year",
       y = "NO2 Concentration (ppb)",
       color = "State") +
  theme(legend.position = "bottom")

# PM10 

#ggplot(df_PM10_top3, aes(x = State, y = Mean, fill = State)) +
#  geom_boxplot() +
#  labs(title = "Distribution of PM10 Levels Across States",
#       x = "State",
#       y = "PM10 Concentration (µg/m³)") +
#  theme_minimal() +
#  theme(legend.position = "none")

# SO2

ggplot(df_SO2_top3, aes(x = Year,
                        y = Mean,
                        color = State)) +
  geom_boxplot() +
  labs(title = "Distribution of SO2 Levels Across States",
       x = "Year",
       y = "SO2 Concentration (ppb)",
       color = "State") +
  theme(legend.position = "bottom")


```
  
### Interpretation of the Boxplots
From the boxplots, we noted the following:

PM2.5 Levels: The distribution of PM2.5 levels varies widely between states. California shows a particularly high range of PM2.5 concentrations with notable outliers, indicating episodes of very poor air quality. It's important to look into the reasons for California's variability, such as wildfires or urban pollution.

NO2 Levels: NO2 levels are somewhat variable, with California showing a higher median concentration. This could be associated with industrial activities or high traffic density.

SO2 Levels: Iowa has historically had a higher range of SO2 concentrations which have fluctuated over time and trended downwards. This pollutant is often associated with industrial processes and the burning of sulfur-containing fuels such as coal. The coal mining and energy generation activity in Iowa could explain the relatively higher concentrations of SO2 measure relative to California and Texas.


\newpage

# Summary and Conclusions  

As shown in the analysis, there has been a general upward trend of renewable energy generation over time in the United States, indicating increased adoption and capacity over time. California (CA) stands out with a significantly higher generation, especially with a steep increase around 2020. Other states also show growth in renewable energy generation but to varying degrees. For instance, Texas (TX) and Iowa (IA) show notable increases. The variability in generation over time could be influenced by factors like state policies, technological advancements, and investment in renewable energy infrastructure.The overall increasing trend aligns with global efforts to transition to cleaner energy sources to reduce reliance on fossil fuels and combat climate change.

Our analysis suggest that the measured value of the pollutants has an inverse relationship or negative correlation with net generation from solar and wind in the three states. This implies that the higher the amount of energy generated from wind and solar, the lower the amount of the three criteria pollutants.



\newpage

# References
Kabeyi, Moses Jeremiah Barasa, and Oludolapo Akanni Olanrewaju. ‘Sustainable Energy Transition for Renewable and Low Carbon Grid Electricity Generation and Supply’. Frontiers in Energy Research, vol. 9, 2022. Frontiers, https://www.frontiersin.org/articles/10.3389/fenrg.2021.743114.

Perera FP. 2017. Multiple threats to child health from fossil fuel combustion: Impacts of air pollution and climate change. Environmental Health Perspectives 125:141-148.

Perera, F., Ashrafi, A., Kinney, P., & Mills, D. (2019). Towards a fuller assessment of benefits to children's health of reducing air pollution and mitigating climate change due to fossil fuel combustion. Environmental research, 172, 55-72.

Bridges, A., Felder, F. A., McKelvey, K., & Niyogi, I. (2015). Uncertainty in energy planning: Estimating the health impacts of air pollution from fossil fuel electricity generation. Energy Research & Social Science, 6, 74-77.

Environmental Impacts of Renewable Energy Technologies | Union of Concerned Scientists. https://www.ucsusa.org/resources/environmental-impacts-renewable-energy-technologies. Accessed 6 Dec. 2023.

Air Pollution. https://education.nationalgeographic.org/resource/air-pollution. Accessed 6 Dec. 2023.
