# EIA-860 (https://www.eia.gov/electricity/data/eia860/)

## Summary
This dataset collects generator-level specific information about existing and planned generators and associated environmental equipment at electric power plants with 1 megawatt or greater of combined nameplate capacity.

## Database Information
Download the zip files for the respective years at https://www.eia.gov/electricity/data/eia860/. 
LayoutYyyyy– Provides a directory of all (published) data elements collected on the Form EIA-860 together with the related description, specific file location(s), and, where appropriate, an explanation of codes.

1___UtilityYyyyy - Contains utility-level data for the plants and generators surveyed in the reporting year.

2___PlantYyyyy - Contains plant-level data for the generators surveyed in all available years.

3_1_GeneratorYyyyy – Contains generator-level data for the surveyed generators, split into three tabs.
The Operable tab includes those generators which are currently operating, out of service or on standby;
The Proposed tab includes those generators which are planned and not yet in operation; and
The Retired and Canceled tab includes those generators which were cancelled prior to completion and operation and retired generators at existing plants.
The retired tab only includes those retired generators which were reported in the most current data cycle. This is not a comprehensive list. Starting with March 2017 data, Preliminary Monthly Electric Generator Inventory includes a comprehensive list of generators which retired since 2002. The list can be found on the Retired tab of the data file.

3_2_WindYyyyy – Contains additional details for surveyed generators that use wind as an energy source, split into two tabs:
The Operable tab includes those generators which are currently operating, out of service or on standby; and
The Retired and Canceled tab includes those generators which were cancelled prior to completion and operation and retired generators at existing plants.

3_3_SolarYyyyy - Contains additional details for surveyed generators that use solar as an energy source, split into two tabs:
The Operable tab includes those generators which are currently operating, out of service or on standby;
The Retired and Canceled tab includes those generators which were cancelled prior to completion and operation and retired generators at existing plants.

3_4_Energy_StorageYyyyy – Contains additional details of surveyed generators for the energy storage technology, split into two tabs:
The Operable tab includes those generators which are currently operating, out of service or on standby;
The Retired and Canceled tab includes those generators which were cancelled prior to completion and operation and retired generators at existing plants.

3_5_MultiFuelYyyyy – Contains data on fuel-switching and the use of multiple fuels by surveyed generators, split into three tabs:
The Operable tab includes those generators which are currently operating, out of service or on standby; and
The Proposed tab includes those generators which are planned and not yet in operation; and
The Retired and Canceled tab includes those generators which were cancelled prior to completion and operation and retired generators at existing plants.

4___OwnerYyyyy – Contains owner and/or operator data for generators with shared ownership and generators that are wholly-owned by an entity other than the operator (generators not appearing in the file are wholly-owned by their operator).

6_1_EnviroAssocYyyyy – Contains boiler association data for the environmental equipment data collected on the Form EIA-860.
The Boiler Generator identifies which boilers are associated with each generator;
The Boiler Cooling tab shows which cooling systems are associated with each boiler;
The Boiler Particulate Matter tab shows which flue gas particulate (FGP) collectors are associated with each boiler;
The Boiler SO2 tab shows which flue gas desulfurization (FGD) systems are associated with each boiler
The Boiler NOx tab shows which nitrogen oxide control equipment is associated with each boiler;
The Boiler Mercury tab shows which mercury control equipment is associated with each boiler
The Boiler Stack Flue tab shows which stacks and flues are associated with each boiler; and
The Emissions Control Equipment tab shows the operational status, in-service date, and installation costs of all the environmental equipment.

6_2_EnviroEquipYyyyy – Contains environmental equipment data for the surveyed generators.
The Emission Standards & Strategies tab shows boiler data as collected on Schedule 6, Part B of the Form EIA-860;
The Boiler Info & Design Parameters tab shows boiler data as collected on Schedule 6, Part C;
The Cooling tab shows cooling system data as collected on Schedule 6, Part D;
The FGP tab shows FGP data as collected on Schedule 6, Part E;
The FGD tab shows FGD data as collected on Schedule 6, Part F; and
The StackFlue tab shows stack and flue data as collected on Schedule 6, Part G.


## Data Content Information
From the downloaded .zip files, only the following two files of interest were retained for the analysis: 
- 2___PlantYyyyy 
- 3_1_GeneratorYyyyy file, only the data in the Operable tab was used for the analysis.

The column names have descriptors hence are self-explanatory.

